Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-04 response.tex

The WPE vs. MASE points of the transformed data move---upwards,
generally---but they remain within the shaded region.  Indeed, the
values for the transformed version of {\tt dgesdd}$_5$ (dark-green
icons) move {\sl closer} to the curve.

This series of experiments supports our claims of generality: a
transformation like this effectively changes the generating process,
but does not affect the performance of our model-free heuristic.  And
the curves that capture that heuristic---which were derived from data
from a single computer-performance experiment---do not only hold for
traces from that experiment and transformed versions of them.  The
preliminary results of our broader study, currently underway, indicate
that they also hold for data from synthetic examples and for
real-world data from a different system.  The image below shows some
of these results:

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras_with_nonlinear_all_points}
\end{center}

The blue fit lines here are derived from the original 120 traces; the
magenta one is fit to those traces plus the nonlinear transformations
of {\tt dgesdd}$_2$, {\tt dgesdd}$_5$, and {\tt dgesdd}$_6$, the $x$
coordinate of a 10,000-point orbit from the H\'{e}non map with $a=1.4$
and $b=0.3$, the $x$ coordinate of a 10,000-step 4th-order Runge-Kutta
solution of the Lorenz equations with $r=28, \; \sigma=10, \; b=8/3$
and $\Delta t=0.01$, a 10,000-step random-walk time series generated by
taking the cumulative sum of $\mu=0$, $\sigma=1$ Gaussian random
variables using Numpy ({\tt numpy.cumsum(numpy.random.normal(0.0, 1.0, size=10000))}),
and data set ``A'' from the Santa Fe Institute time-series competition
website.

### 2014-08-04 intro.tex
few more comments

successfully predicted from past values.  Signals like this can be

@@

midrange of this spectrum---e.g., deterministic chaos---pose
interesting challenges from a modeling perspective.  In these signals,
enough information is being transmitted from the past to the future
that an \emph{ideal} model---one that captures the generating
process---can forecast the future behavior of the observed system with
high accuracy.

@@

This leads naturally to an important and challenging question: given a
noisy real-valued time series from an unknown system, does there exist
a forecast model that can leverage the information (if any) that is
being transmitted from past to future by the underlying generating
process?  A first step in answering this question is to reliably
quantify where on the complexity spectrum a given time series falls; a
second step is to determine how complexity and predictability are
related in these kinds of data sets.  With these answers in hand, one
can develop a practical strategy for assessing appropriateness of
forecast methods for a given time series.  If the forecast produced by
a particular method is poor, for instance, but the time series
contains a significant amount of predictive structure, one can
reasonably conclude that that method is inadequate to the task and
that one should seek another method.  The goal of this paper is to
develop effective heuristics to put that strategy into practice.

@@

We focus on real-valued, scalar, time-series data from physical experiments.
%
% that appears complicated: i.e., not periodic, linear, constant, etc.
%
We do not assume any knowledge of the generating process or its
properties: whether it is linear, nonlinear, deterministic,
stochastic, stationary, non-stationary, etc.  To explore the
relationship between complexity, predictive structure, and
predictability, we generate forecasts for a variety of experimental
time-series datasets using four different prediction methods, then
compare the accuracy of those predictions to the permutation entropy
of the associated signals.  This results in two primary findings:
% 
\begin{enumerate}
% \item The complexity of a noisy real-valued time series can be
%   effectively quantifiable by computing its permutation entropy.

>>>>

successfully\alert{perfectly?} predicted from past values.  Signals like this can be

@@

midrange of this spectrum, e.g., deterministic chaos, pose interesting
challenges from a modeling perspective.  In these signals, enough
information is being transmitted from the past to the future that an
\emph{ideal} model---one that captures the generating process---can
forecast the future behavior of the observed system with high
accuracy.

@@

This leads naturally to an important and challenging question: given a
noisy real-valued time series from an unknown system, does there exist
a forecast model that can leverage the information (if any) that is
being transmitted forward in time by the underlying generating
process?  A first step in answering this question is to reliably
quantify where on the complexity spectrum a given time series falls; a
second step is to determine how complexity and predictability are
related in these kinds of data sets.  With these answers in hand, one
can develop a practical strategy for assessing appropriateness of
forecast methods for a given time series.  If the forecast produced by
a particular method is poor, for instance, but the time series
contains a significant amount of predictive structure, one can
reasonably conclude that that method is inadequate to the task and
that one should seek another method.  The goal of this paper is to
develop effective heuristics to put that strategy into practice.

@@

We focus on real-valued, scalar, time-series data from physical experiments.
%
% that appears complicated: i.e., not periodic, linear, constant, etc.
%
We do not assume any knowledge of the generating process or its
properties: whether it is linear, nonlinear, deterministic,
stochastic, etc.  To explore the relationship between complexity,
predictive structure, and actual predictability, we generate forecasts
for a variety of experimental time-series datasets using four
different prediction methods, then compare the accuracy of those
predictions to the permutation entropy of the associated signals.
This results in two primary findings:
% 
\begin{enumerate}
% \item The complexity of a noisy real-valued time series can be
%   effectively quantifiable by computing its permutation entropy.
